---
title: "Problem Set 7"
author: "Kerri Rose Riley"
format:
  html: default
  pdf: default
---

## 1. Replicate both parts of Figure 1 for Bartel
```{r}
# Import modules
library(tidyverse) # For data wrangling
library(haven) # To use read_dta() for Data files
library(ggplot2) # For graphs and plots

# Read the ANES data
anes_2024 <- read_dta("~/Desktop/Current Classes/R/Problem Set 7/Data/anes_2024.dta")

# Read the cumulative ANES data
anes_cum <- read_dta("~/Desktop/Current Classes/R/Problem Set 7/Data/anes_cum.dta")


# Recode VCF0305 into four party identification categories

figure1 <- anes_cum |>
  # Filter to have same years
  filter(VCF0004 >= 1952 & VCF0004 <= 1996) |>
  mutate(
    party_id = case_when(
      VCF0305 == 1 ~ "Pure Independents",
      VCF0305 == 2 ~ "Independent Leaners",
      VCF0305 == 3 ~ "Weak Identifiers",
      VCF0305 == 4 ~ "Strong Identifiers",
      TRUE ~ NA_character_
    )
  ) |>
  filter(!is.na(party_id))

# Calculate proportions
figure1 <- figure1 |>
  group_by(VCF0004, party_id) |>
  summarize(n = n(), .groups = "drop") |>
  group_by(VCF0004) |>
  mutate(proportion = n / sum(n)) |>
  ungroup()


figure1 <- figure1 |>
  mutate(
    panel = case_when(
      party_id %in% c("Strong Identifiers", "Weak Identifiers") ~ "Identifiers",
      party_id %in% c("Independent Leaners", "Pure Independents") ~ "Independents"
    )
  )


# PLOTTING
ggplot(figure1, aes(
  x = VCF0004,
  y = proportion,
  group = party_id,
  linetype = party_id,
  shape = party_id
)) +
  # Lines connect all years
  geom_line() +
  
  # Points only at labeled years
  geom_point(
    data = filter(
      figure1,
      VCF0004 %% 4 == 0
    ),
    size = 2
  ) +
  
  facet_grid(panel ~ ., scales = "fixed") +
  
  scale_y_continuous(
    limits = c(0, 0.5),
    breaks = seq(0.1, 0.5, 0.1)
  ) +
  
  scale_x_continuous(
    breaks = c(1956, 1964, 1972, 1980, 1988, 1996)
  ) +
  
  scale_linetype_manual(
    values = c(
      "Strong Identifiers" = "solid",
      "Weak Identifiers" = "dashed",
      "Independent Leaners" = "dashed",
      "Pure Independents" = "solid"
    )
  ) +
  
  scale_shape_manual(
    values = c(
      "Strong Identifiers" = 16,
      "Weak Identifiers" = 1,
      "Independent Leaners" = 1,
      "Pure Independents" = 16
    )
  ) +
  
  labs(
    x = NULL,
    y = NULL,
    title = "Proportions of National Election Study Sample"
  ) +
  
  theme_classic() +
  
  theme(
    legend.position = "bottom",
    strip.background = element_blank(),
    strip.text = element_blank(),
    panel.border = element_rect(
      color = "black",
      fill = NA
    ),
    plot.title = element_text(
      hjust = 0.5
    )
  )
```


## 3. Extend the analysis to 2024, using the 2024 ANES file. (This will require using the 2024 file in conjunction with the cumulative file.)
```{r}
# Create the same four party identification categories
# for the 2024 ANES data

anes_2024 <- anes_2024 |>
  mutate(
    VCF0004 = 2024,
    party_id = case_when(
      
      # Pure Independents
      V241221 == 3 & V241223 == 2 ~ "Pure Independents",
      
      # Independent Leaners
      V241221 == 3 & V241223 %in% c(1, 3) ~ "Independent Leaners",
      
      # Strong Identifiers
      V241221 %in% c(1, 2) & V241222 == 1 ~ "Strong Identifiers",
      
      # Weak Identifiers
      V241221 %in% c(1, 2) & V241222 == 2 ~ "Weak Identifiers",
      
      TRUE ~ NA_character_
    )
  )

# Create the same party_id variable for the cumulative data
anes_cum_extended <- anes_cum |>
  mutate(
    party_id = case_when(
      VCF0305 == 1 ~ "Pure Independents",
      VCF0305 == 2 ~ "Independent Leaners",
      VCF0305 == 3 ~ "Weak Identifiers",
      VCF0305 == 4 ~ "Strong Identifiers",
      TRUE ~ NA_character_
    )
  )

# Combine cumulative ANES and 2024 ANES
anes_all <- bind_rows(
  anes_cum_extended,
  anes_2024
)

# Calculate proportions
figure1_extended <- anes_all |>
  filter(VCF0004 >= 1952 & VCF0004 <= 2024) |>
  filter(!is.na(party_id)) |>
  group_by(VCF0004, party_id) |>
  summarize(
    n = n(),
    .groups = "drop"
  ) |>
  group_by(VCF0004) |>
  mutate(
    proportion = n / sum(n)
  ) |>
  ungroup() |>
  mutate(
    panel = case_when(
      party_id %in% c(
        "Strong Identifiers",
        "Weak Identifiers"
      ) ~ "Identifiers",
      
      party_id %in% c(
        "Independent Leaners",
        "Pure Independents"
      ) ~ "Independents"
    )
  )

ggplot(
  figure1_extended,
  aes(
    x = VCF0004,
    y = proportion,
    group = party_id,
    linetype = party_id,
    shape = party_id
  )
) +
  geom_line() +
  geom_point(
    data = filter(figure1_extended, VCF0004 %% 4 == 0),
    size = 2
  ) +
  facet_grid(panel ~ ., scales = "fixed") +
  scale_y_continuous(
    limits = c(0, 0.5),
    breaks = seq(0.1, 0.5, 0.1)
  ) +
  scale_x_continuous(
    breaks = seq(1952, 2024, 8)
  ) +
  scale_linetype_manual(
    values = c(
      "Strong Identifiers" = "solid",
      "Weak Identifiers" = "dashed",
      "Independent Leaners" = "dashed",
      "Pure Independents" = "solid"
    )
  ) +
  scale_shape_manual(
    values = c(
      "Strong Identifiers" = 16,
      "Weak Identifiers" = 1,
      "Independent Leaners" = 1,
      "Pure Independents" = 16
    )
  ) +
  labs(
    x = NULL,
    y = NULL,
    title = "Proportions of National Election Study Sample"
  ) +
  theme_classic() +
  theme(
    legend.position = "bottom",
    strip.background = element_blank(),
    strip.text = element_blank(),
    panel.border = element_rect(
      color = "black",
      fill = NA
    ),
    plot.title = element_text(hjust = 0.5)
  )
```


## 4. Create figure that include midterm election years as well
```{r}
# Create figure including presidential and midterm election years
# Use the cumulative ANES data

 
figure1_midterms <- anes_cum |>
  filter(VCF0004 >= 1952 & VCF0004 <= 2020) |>
  mutate(
    party_id = case_when(
      VCF0305 == 1 ~ "Pure Independents",
      VCF0305 == 2 ~ "Independent Leaners",
      VCF0305 == 3 ~ "Weak Identifiers",
      VCF0305 == 4 ~ "Strong Identifiers",
      TRUE ~ NA_character_
    )
  ) |>
  filter(!is.na(party_id)) |>
  group_by(VCF0004, party_id) |>
  summarize(n = n(), .groups = "drop") |>
  group_by(VCF0004) |>
  mutate(proportion = n / sum(n)) |>
  ungroup() |>
  mutate(
    panel = case_when(
      party_id %in% c("Strong Identifiers", "Weak Identifiers") ~ "Identifiers",
      party_id %in% c("Independent Leaners", "Pure Independents") ~ "Independents"
    )
  )

# Plot
ggplot(
  figure1_midterms,
  aes(
    x = VCF0004,
    y = proportion,
    group = party_id,
    linetype = party_id,
    shape = party_id
  )
) +
  
  # Lines connect all available ANES years
  geom_line() +
  
  # Points show all available ANES years, including midterms
  geom_point(size = 2) +
  
  # Create two panels
  facet_grid(
    panel ~ .,
    scales = "fixed"
  ) +
  
  # Y-axis
  scale_y_continuous(
    limits = c(0, 0.5),
    breaks = seq(0.1, 0.5, 0.1)
  ) +
  
  # X-axis
  scale_x_continuous(
    breaks = seq(1952, 2020, 4)
  ) +
  
  # Line types
  scale_linetype_manual(
    values = c(
      "Strong Identifiers" = "solid",
      "Weak Identifiers" = "dashed",
      "Independent Leaners" = "dashed",
      "Pure Independents" = "solid"
    )
  ) +
  
  # Point shapes
  scale_shape_manual(
    values = c(
      "Strong Identifiers" = 16,
      "Weak Identifiers" = 1,
      "Independent Leaners" = 1,
      "Pure Independents" = 16
    )
  ) +
  
  # Labels
  labs(
    x = NULL,
    y = NULL,
    title = "Proportions of National Election Study Sample"
  ) +
  
  # Theme
  theme_classic() +
  
  theme(
    legend.position = "bottom",
    strip.background = element_blank(),
    strip.text = element_blank(),
    
    # Border around each panel
    panel.border = element_rect(
      color = "black",
      fill = NA
    ),
    
    # Center title
    plot.title = element_text(
      hjust = 0.5
    )
  )
```


## 5. Original Work

```{r} 
political_info <- anes_cum |>
  mutate(
    political_info = case_when(
      VCF0050b == 1 ~ "Very high",
      VCF0050b == 2 ~ "Fairly high",
      VCF0050b == 3 ~ "Average",
      VCF0050b == 4 ~ "Fairly low",
      VCF0050b == 5 ~ "Very low",
      TRUE ~ NA_character_
    )
  ) |>
  filter(!is.na(political_info)) |>
  mutate(
    political_info = factor(
      political_info,
      levels = c(
        "Very high",
        "Fairly high",
        "Average",
        "Fairly low",
        "Very low"
      )
    )
  )
# Plot
ggplot(political_info, aes(x = political_info)) +
  geom_bar() +
  labs(
    x = "Level of Political Information",
    y = "Number of Respondents"
  ) +
  theme_classic()


# VCF0504 - the republicans party liberal to conservative scale
republican_ideology <- anes_cum |>
  mutate(
    republican_ideology = case_when(
      VCF0504 %in% c(1, 2, 3) ~ "Liberal",
      VCF0504 == 4 ~ "Moderate",
      VCF0504 %in% c(5, 6, 7) ~ "Conservative",
      TRUE ~ NA_character_
    )
  ) |>
  filter(!is.na(republican_ideology)) |>
  mutate(
    republican_ideology = factor(
      republican_ideology,
      levels = c(
        "Liberal",
        "Moderate",
        "Conservative"
      )
    )
  )

ggplot(republican_ideology, aes(x = republican_ideology)) +
  geom_bar() +
  labs(
    x = "Rating of Republican Party",
    y = "Number of Respondents"
  ) +
  theme_classic()

```


## Follow-up Questions
### Where did you encounter problems with this replication? Was there anything that was particularly tricky?

#### The most difficult part was making sure the party identification categories were coded consistently with the original figure, especially when working with the different variables in the cumulative ANES file and the 2024 ANES file. It was also tricky to combine the 2024 data with the cumulative data because the variables used to measure party identification are different between the two files.


###How would you interpret (explain) what Bartels found in figures 1 & 2?

#### Bartels’ Figure 1 shows a clear decline in strong party identification and an increase in independent identification between 1952 and 1996. The important point is that this does not mean Americans simply stopped having partisan preferences; much of the movement was toward weaker forms of party identification, especially independent leaners. In the top panel, the share of strong party identifiers was around 35% in the 1950s and early 1960s, but fell sharply after 1964, reaching roughly 23–25% in the 1970s. It recovered somewhat by the 1990s but remained below its earlier level. In the bottom panel, independent leaners rose from roughly 15% in the 1950s to about 25–27% by the late 1980s and early 1990s.


###Has anything changed since 1996 that would alter the substantive takeaway points in their findings?

#### After 1996, the proportion of “Strong” Identifiers generally increased, while the proportion of “Weak” Identifiers decreased. At the same time, both “Pure” Independents and Independent “Leaners” generally declined. This suggests a reversal of the trend Bartels observed in the earlier period: after 1996, Americans appear to have become more strongly attached to political parties, while fewer identified as independents.


###Does anything change about the analysis in a substantive way when you include midterm election years?

#### 


