---
title: "Problem Set 7"
author: "Kerri Rose Riley"
format:
  html: default
  pdf: default
---

## 1. Replicate both parts of Figure 1 for Bartel
```{r}
# Import modules
library(tidyverse) # For data wrangling
library(haven) # To use read_dta() for Stata files
library(ggplot2) # For graphs and plots

# Read the ANES data
anes_2024 <- read_dta("~/Desktop/Current Classes/R/Problem Set 7/Data/anes_2024.dta")

# Read the cumulative ANES data
anes_cum <- read_dta("~/Desktop/Current Classes/R/Problem Set 7/Data/anes_cum.dta")


# Recode VCF0305 into four party identification categories

figure1 <- anes_cum |>
  # Filter to have same years
  filter(VCF0004 >= 1952 & VCF0004 <= 1996) |>
  mutate(
    party_id = case_when(
      VCF0305 == 1 ~ "Pure Independents",
      VCF0305 == 2 ~ "Independent Leaners",
      VCF0305 == 3 ~ "Weak Identifiers",
      VCF0305 == 4 ~ "Strong Identifiers",
      TRUE ~ NA_character_
    )
  ) |>
  filter(!is.na(party_id))

# Calculate proportions
figure1 <- figure1 |>
  group_by(VCF0004, party_id) |>
  summarize(n = n(), .groups = "drop") |>
  group_by(VCF0004) |>
  mutate(proportion = n / sum(n)) |>
  ungroup()


figure1 <- figure1 |>
  mutate(
    panel = case_when(
      party_id %in% c("Strong Identifiers", "Weak Identifiers") ~ "Identifiers",
      party_id %in% c("Independent Leaners", "Pure Independents") ~ "Independents"
    )
  )


# PLOTTING
ggplot(figure1, aes(
  x = VCF0004,
  y = proportion,
  group = party_id,
  linetype = party_id,
  shape = party_id
)) +
  # Lines connect all years
  geom_line() +
  
  # Points only at labeled years
  geom_point(
    data = filter(
      figure1,
      VCF0004 %% 4 == 0
    ),
    size = 2
  ) +
  
  facet_grid(panel ~ ., scales = "fixed") +
  
  scale_y_continuous(
    limits = c(0, 0.5),
    breaks = seq(0.1, 0.5, 0.1)
  ) +
  
  scale_x_continuous(
    breaks = c(1956, 1964, 1972, 1980, 1988, 1996)
  ) +
  
  scale_linetype_manual(
    values = c(
      "Strong Identifiers" = "solid",
      "Weak Identifiers" = "dashed",
      "Independent Leaners" = "dashed",
      "Pure Independents" = "solid"
    )
  ) +
  
  scale_shape_manual(
    values = c(
      "Strong Identifiers" = 16,
      "Weak Identifiers" = 1,
      "Independent Leaners" = 1,
      "Pure Independents" = 16
    )
  ) +
  
  labs(
    x = NULL,
    y = NULL,
    title = "Proportions of National Election Study Sample"
  ) +
  
  theme_classic() +
  
  theme(
    legend.position = "bottom",
    strip.background = element_blank(),
    strip.text = element_blank(),
    panel.border = element_rect(
      color = "black",
      fill = NA
    ),
    plot.title = element_text(
      hjust = 0.5
    )
  )
```

## 2. Extend Analysis to include recent years
```{r}

```

## 3.Extend the analysis to 2024, using the 2024 ANES file. (This will require using the 2024 file in conjunction with the cumulative file.)
```{r}

```


## 4. Create figure that include midterm election years as well
```{r}

```


## 5. Original Work

```{r}

```


## Follow-up Questions
### Where did you encounter problems with this replication? Was there anything that was particularly tricky?
####

###How would you interpret (explain) what Bartels found in figures 1 & 2?
####

###Has anything changed since 1996 that would alter the substantive takeaway points in their findings?
####

###Does anything change about the analysis in a substantive way when you include midterm election years?
####


